---
title: "METHOD 3"
author: "Jüttner, D., Labarthe, S., and Prastalo, N."
date: "2026"
output:
  pdf_document: default
  html_document: default
---

# 1. CODEBOOK
* `iso3c` - ISO 3-letter country code
* `year` - Tournament year
* `v2exl_legitideolcr_0_1` - Nationalism (1 year lead)
* `v2exl_legitideolcr_0_3` - Nationalism (3 year lead)
* `treatment` - World Cup qualification treatment
* `points_gap` - Points gap between last qualifier and first non-qualifier
* `points_diff` - Assignment variable (points margin from cutoff)
* `wdi_expmil` - Military expenditure (% of GDP)
* `wdi_pop_log` - Total population (log)
* `wdi_popurb` - Urban population
* `wdi_lifexp` - Life expectancy at birth
* `wdi_export` - Exports of goods and services (% of GDP)
* `wdi_import` - Imports of goods and services (% of GDP)
* `wdi_gdpcapcon2015` - GDP per capita (constant 2015 US$)
* `wdi_gdpgr` - GDP growth (annual %)
* `wdi_area` - Land area (sq. km)
* `wpp_medianage` - Median age of population
* `wpp_sexratio` - Sex ratio (males per 100 females)
* `vdem_polyarchy` - Polyarchy (electoral democracy index)

# 2. DATABASE REPLICATION
```{r, eval = F}
## a. VDEM 
library(dplyr)
library(countrycode)
library(countries)
library(tidyr)
library(readr)

vdem <- read_rds("https://raw.githubusercontent.com/SimonLabarthe/Method-3/main/vdem.rds") %>%  
  mutate(v2exl_legitideolcr_0_1 = lead(v2exl_legitideolcr_0, 1), 
         v2exl_legitideolcr_0_3 = lead(v2exl_legitideolcr_0, 3)) 

vdem_clean <- vdem %>%
  mutate(iso3c = country_name(vdem$country_name, 
                              to = "ISO3", 
                              fuzzy_match = TRUE,
                              verbose = FALSE)) %>%
  filter(!is.na(iso3c))

## b. QOG
url_data <- "https://raw.githubusercontent.com/SimonLabarthe/Method-3/main/qog.rds"
qog <- read_rds(url_data) %>% 
  rename(country_name = cname) %>% 
  select(
    country_name,
    year,

    wdi_expmil,         # Military expenditure % of GDP (i = 163, t=1960-2024))

    wdi_pop,            # Total Population (i = 192, t = 1960-2024)
    wdi_popurb,         # Urban Population (i = 192, t = 1960-2024)
    wdi_lifexp,         # Life Expectancy at birth (per 1,000 live births) (i = 192, t = 1960-2023)

    wpp_medianage,      # Median Age as of 1 July (WPP) (i = 192, t = 1950-2024)
    wpp_sexratio,       # Sex Ratio: males per 100 females (WPP) (i = 192, t = 1950-2024)


    wdi_export,         # Exports of goods & services (% of GDP) (i = 173, t = 1960-2024)
    wdi_import,         # Imports of goods & services (% of GDP) (i = 173, t = 1960-2024)
    wdi_gdpcapcon2015,  # GDP per capita (constant 2015 US dollar) (i = 191, t = 1960-2024)
    wdi_gdpgr,          # GDP growth (annual %) (i = 191, t = 1961-2024)

    wdi_area,           # Land area in sq. km (i = 191, t = 1961-2023)

    vdem_polyarchy     # Democracy level (i = 175, t = 1946-2024)
  ) %>% 
  mutate(
    wdi_pop_log = log(wdi_pop)
  )

qog_clean <- qog %>%
  mutate(iso3c = country_name(country_name, # here again
                             to = "ISO3", 
                             fuzzy_match = TRUE,
                             verbose = FALSE)) %>%
  filter(!is.na(iso3c))

## c. FIFA World Cup
library(tidyverse)
library(worldcup)
library(lubridate)

data("matches")
data("tournaments")
data("teams")
data("qualified_teams")

qualification_matches <- matches %>%
  filter(stage_name == "group stage") %>%
  mutate(
    year = as.numeric(str_extract(tournament_name, "\\d{4}"))
  ) %>%
  filter(year %% 2 == 0) %>% # males only
  mutate(
    home_team_name = if_else(home_team_name == "England", "United Kingdom", home_team_name), # England as United Kingdom
    home_iso3c = country_name(home_team_name, to = "ISO3", fuzzy_match = TRUE, verbose = FALSE), 
    
    away_team_name = if_else(away_team_name == "England", "United Kingdom", away_team_name),
    away_iso3c = country_name(away_team_name, to = "ISO3", fuzzy_match = TRUE, verbose = FALSE) 
  ) %>%
  filter(!is.na(home_iso3c), !is.na(away_iso3c))

home_team_matches <- qualification_matches %>%
  select(
    tournament_id, 
    iso3c = home_iso3c,  
    year, match_id, match_date,
    group_name, group_stage, stage_name,
    country_name = home_team_name,
    opponent_name = away_team_name,
    goals_for = home_team_score,
    goals_against = away_team_score,
    win = home_team_win, draw, loss = away_team_win
  ) %>%
  mutate(points = case_when(win == 1 ~ 3, draw == 1 ~ 1, TRUE ~ 0),
         goal_difference = goals_for - goals_against,
         venue = "home")

away_team_matches <- qualification_matches %>%
  select(
    tournament_id,
    iso3c = away_iso3c,  
    year, match_id, match_date,
    group_name, group_stage, stage_name,
    country_name = away_team_name,
    opponent_name = home_team_name,
    goals_for = away_team_score,
    goals_against = home_team_score,
    win = away_team_win, draw, loss = home_team_win
  ) %>%
  mutate(points = case_when(win == 1 ~ 3, draw == 1 ~ 1, TRUE ~ 0),
         goal_difference = goals_for - goals_against,
         venue = "away")

monadic_matches <- bind_rows(home_team_matches, away_team_matches)

qualification_standings <- monadic_matches %>%
  group_by(year, group_name, iso3c) %>%
  summarise(
    matches_played = n(),
    wins = sum(win, na.rm = TRUE),
    draws = sum(draw, na.rm = TRUE),
    losses = sum(loss, na.rm = TRUE),
    goals_for = sum(goals_for, na.rm = TRUE),
    goals_against = sum(goals_against, na.rm = TRUE),
    goal_difference = sum(goal_difference, na.rm = TRUE),
    points = sum(points, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(year, group_name, desc(points), desc(goal_difference), desc(goals_for))

qualification_standings <- qualification_standings %>%
  group_by(year, group_name) %>%
  mutate(
    rank = row_number(),
    teams_qualifying = if_else(year <= 1970, 1, 2), # number of qualified teams
    qualified = if_else(rank <= teams_qualifying, 1, 0)
  ) %>%
  ungroup()

# filter : keep only tight races
tight_races <- qualification_standings %>%
  group_by(year, group_name) %>%
  mutate(
    min_qualifier_points = min(points[qualified == 1], na.rm = TRUE),
    max_non_qualifier_points = max(points[qualified == 0], na.rm = TRUE),
    points_gap = min_qualifier_points - max_non_qualifier_points
  ) %>%
  ungroup() %>%
  filter((qualified == 1 & points == min_qualifier_points) |
         (qualified == 0 & points == max_non_qualifier_points)) %>%
  mutate(
    treatment = qualified,
    points_diff = if_else(qualified == 1, points_gap, -points_gap)
  )

## d. Merge
data <- tight_races %>%
  left_join(qog_clean, by = c("iso3c", "year")) %>% 
  left_join(vdem_clean, by = c("iso3c", "year")) %>% 
  filter(!is.na(treatment), 
         points_diff != "Inf") 

data <- data %>% 
  select(iso3c, year, v2exl_legitideolcr_0_1, v2exl_legitideolcr_0_3,  treatment, points_gap, points_diff, wdi_expmil, wdi_pop_log, wdi_popurb, wdi_lifexp, wdi_export, wdi_import, wdi_gdpcapcon2015, wdi_gdpgr, wdi_area, wpp_medianage, wpp_sexratio, vdem_polyarchy) %>% na.omit()

library(labelled)

var_labels <- c(
  "iso3c" = "ISO 3-letter country code",
  "year" = "Tournament year",
  "v2exl_legitideolcr_0_1" = "Nationalism (1 year lead)",
  "v2exl_legitideolcr_0_3" = "Nationalism (3 year lead)",
  "treatment" = "Treatment: World Cup qualification (1=qualified, 0=non-qualified)",
  "points_gap" = "Points gap between last qualifier and first non-qualifier",
  "points_diff" = "Assignment variable: points margin from cutoff (positive=qualified)",
  "wdi_expmil" = "Military expenditure (% of GDP)",
  "wdi_pop_log" = "Total population (log)",
  "wdi_popurb" = "Urban population",
  "wdi_lifexp" = "Life expectancy at birth",
  "wdi_export" = "Exports of goods and services (% of GDP)",
  "wdi_import" = "Imports of goods and services (% of GDP)",
  "wdi_gdpcapcon2015" = "GDP per capita (constant 2015 US$)",
  "wdi_gdpgr" = "GDP growth (annual %)",
  "wdi_area" = "Land area (sq. km)",
  "wpp_medianage" = "Median age of population",
  "wpp_sexratio" = "Sex ratio (males per 100 females)",
  "vdem_polyarchy" = "Polyarchy (electoral democracy index)"
)

data <- labelled::set_variable_labels(data, .labels = var_labels)

look_for(data)

test <- data %>% 
  group_by(iso3c, year) %>% 
  slice(1) %>% 
  na.omit()

saveRDS(data, "data.rds")
```

# 3. LOAD DATA IN CLEAN ENVIRONMENT

```{r}
library(dplyr)
library(readr)
library(ggplot2)

rm(list=ls())

url_data <- "https://raw.githubusercontent.com/SimonLabarthe/Method-3/main/data.rds"

data <- read_rds(url_data) %>% 
  mutate(y1 = v2exl_legitideolcr_0_1, 
         y3 = v2exl_legitideolcr_0_3, 
         d = treatment, 
         z = points_diff) %>%
  select(iso3c, 
         year, 
         y1, 
         y3, 
         d, 
         z, 
         everything(), 
         -v2exl_legitideolcr_0_1, 
         -v2exl_legitideolcr_0_3, 
         -treatment, 
         -points_diff)

x <- data[, 8:19]
```

# 4. ASSUMPTIONS CHECK
## a. Covariates Visualization P-Value
```{r}
library(ggplot2)
library(gridExtra)

data_c0 <- data %>% filter(data$z == 0) # on the cutoff
data_bw1 <- data %>% filter(data$z >= -1 & data$z <= 1) # around the cutoff

# datasets to test
datasets <- c("data", "data_bw1", "data_c0") # multiple datasets possible

# Covariates
vars <- names(x)

# add labs
better_names <- sapply(vars, function(var) {
  label <- attr(data[[var]], "label")
  if(is.null(label)) var else label  
})

plot_list <- list()

# plot the p-values
for(d in datasets) {
  data_current <- get(d)
  balance_p <- rep(NA, length(vars))

  for(i in seq_along(vars)){
    temp_mod <- lm(
      as.formula(paste(vars[i], "~ d")),
      data = data_current[, c(vars[i], "d")]
    )
    balance_p[i] <- summary(temp_mod)$coefficients[2, 4]
  }
  
  # new dataframe for graph
  df <- data.frame(vars = vars, balance_p = balance_p, better_names = better_names)
  df$y <- rank(-df$balance_p, ties.method = "first")
  
  # extract p-values
  cat("\n========== Dataset:", d, "==========\n")
  print(data.frame(Variable = better_names, P_value = round(balance_p, 4)))
  
  # graph
  p <- ggplot(df, aes(x = balance_p, y = y)) +
    geom_point(size = 2, color = "darkblue") +
    scale_x_continuous(expand = c(0,0), limit = c(-0.01, 1.01), 
                       breaks = c(0.05, 0.1, 0.25, 0.5, 0.75, 1)) +
    xlab(expression(paste(italic("p"), "-value", sep = ""))) +
    scale_y_continuous(breaks = df$y, labels = df$better_names) +
    ylab(NULL) +
    geom_vline(xintercept = 0.05, lty = 2, color = "red", linewidth = 0.8) +
    geom_vline(xintercept = 0.10, lty = 2, color = "orange", linewidth = 0.8) +
    theme_bw() +
    theme(panel.grid.major.x = element_blank(),
          panel.grid.minor = element_blank(),
          panel.grid.major.y = element_line(linewidth = 0.15, linetype = "dashed"))
  
  print(p)
  plot_list[[d]] <- p
  ggsave(paste0("balance_plot_", d, ".png"), p, width = 10, height = 6)
}
```

## b. Covariates Visualization Mean Comparison

```{r}
# datasets to test
datasets <- c("data", "data_bw1", "data_c0") # multiple datasets possible

# Covariates
balance_results <- list()

# plot the substantial differences
for (d in datasets) { 
  data_current <- get(d)
  
  balance_table <- data.frame(
    variable = better_names,
    control_mean = NA,
    treatment_mean = NA
    )

  for (i in seq_along(vars)) {
    v <- vars[i]
    balance_table$treatment_mean[i] <- mean(data_current[[v]][data_current$d == 1], na.rm = TRUE)
    balance_table$control_mean[i] <- mean(data_current[[v]][data_current$d == 0], na.rm = TRUE)
  }

  balance_table <- balance_table %>%
    mutate(
      difference = treatment_mean - control_mean,
      treatment_mean = round(treatment_mean, 2),
      control_mean = round(control_mean, 2),
      difference = round(difference, 2)
    )

    balance_results[[d]] <- balance_table
} 

balance_results[["data"]]
balance_results[["data_bw1"]]
balance_results[["data_c0"]]
```

## c. Manipulation

```{r}
manipulation <- ggplot(data, aes(x = z)) +
  geom_histogram(binwidth = 1,
                 color = "black",
                 fill = "darkgrey") +
  geom_density(aes(y = ..density.. * nrow(data) * 0.5),
               color = "darkblue",
               linewidth = 1) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  scale_y_continuous(
    name = "Frequency",
    sec.axis = sec_axis(~ . / (nrow(data) * 0.5),
                        name = "Density")) +
  scale_x_continuous(
    breaks = seq(min(data$z, na.rm = TRUE),
                 max(data$z, na.rm = TRUE),
                 by = 1)) +
  labs(x = "Running variable (margin)") +
  theme_bw()

manipulation
ggsave("manipulation.png", manipulation, width = 10, height = 6)

# McCrary Test------------------------------------------------------------------

#install.packages("rddensity")
library(rddensity)

result <- rddensity(X = data$z, c = 0)
summary(result)
```

# 5. RDD MODELS

## a. Run

```{r}
library(rdrobust)

# Variable "z" is our running variable
model_h5 <- rdrobust(y = data$y1,
         x = data$z,
         c = 0,
         p = 1, 
         kernel = "tri",
         h = 5,
         masspoints = "adjust")

summary(model_h5)

model_mserd <- rdrobust(y = data$y1,
         x = data$z,
         c = 0,
         p = 0, 
         kernel = "tri",
         bwselect = "mserd",
         masspoints = "adjust")

summary(model_mserd)

model_cerrd <- rdrobust(y = data$y1, 
                        x = data$z, 
                        c = 0, 
                        p = 0, 
                        kernel = "tri", 
                        bwselect = "cerrd", 
                        masspoints = "adjust")

summary(model_cerrd)
```

## b. Extract data and print table

```{r}
library(rdrobust)
library(stargazer)
library(webshot)

# model selection
models <- list(
  h5 = model_h5,
  mserd = model_mserd,
  cerrd = model_cerrd
)

# extract the values
extract_rdd_results <- function(model, model_name) {
data.frame(
    Model = model_name,
    Estimate = model$coef[1],           # Robust estimate
    SE = model$se[3],                   # Robust SE
    CI_lower = model$ci[3,1],           # Robust CI lower
    CI_upper = model$ci[3,2],           # Robust CI upper
    p_value = model$pv[3],              # Robust p-value
    Eff_N_total = sum(model$N_h),       # Total effective N
    Bandwidth = model$bws[1,1],         # Bandwidth
    stringsAsFactors = FALSE
  )
}

results_list <- lapply(names(models), function(name) {
  extract_rdd_results(models[[name]], name)
})

# create the table
results_table <- do.call(rbind, results_list)
results_table_formatted <- results_table %>%
  mutate(
    Estimate = sprintf("%.4f", Estimate),
    CI_95 = sprintf("[%.4f, %.4f]", CI_lower, CI_upper),
    `p-value` = sprintf("%.4f", p_value),
    `Eff. N` = sprintf("%d", Eff_N_total),
    `Bandwidth` = sprintf("%.2f", Bandwidth)
  ) %>%
  select(Model, Estimate, CI_95, `p-value`, `Eff. N`, `Bandwidth`)

print(results_table_formatted)

# LaTeX visualisation
stargazer(results_table_formatted, 
                    summary = FALSE, 
          rownames = FALSE, 
          type = "latex",
          out = "rdd_table.tex")
```

# 6. RDD VISUALISATION

```{r}
library(ggplot2)
library(dplyr)

# visualisation
p <- ggplot(data, aes(x = z,
                       y = y1,       
                       color = factor(d))) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = TRUE) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  scale_x_continuous(
    breaks = seq(min(data$z, na.rm = TRUE),
                 max(data$z, na.rm = TRUE), by = 1)
  ) +
  scale_color_manual(
    values = c("0" = "red", "1" = "darkblue"),
    labels = c("Non-qualifier", "Qualifier"),
    name   = "Status"
  ) +
  labs(
    x     = "Points Margin",
    y     = "Nationalism (t+1)" 
  ) +
  theme_bw() +
  theme(legend.position = "bottom")

print(p)

ggsave("rdd_plot_y1.png", p, width = 10, height = 6)
```

# 7. ROBUSTNESS CHECK WITHOUT c = 0 CASES

## a. Refit without the c = 0

```{r}
library(rdrobust)

data_null <- data %>% 
  filter(points_gap != 0)

nrow(data_null) # 132

attach(data_null)

model_h5_null <- rdrobust(y = y1,
         x = z,
         c = 0,
         p = 1, 
         kernel = "tri",
         h = 5,
         masspoints = "adjust")

summary(model_h5_null)
```

## b. Extract data and print table

```{r}
library(rdrobust)
library(stargazer)
library(webshot)

models <- list(
  h5 = model_h5_null
)

extract_rdd_results <- function(model, model_name) {
data.frame(
    Model = model_name,
    Estimate = model$coef[1],           # Estimate
    SE = model$se[3],                   # Robust SE
    CI_lower = model$ci[3,1],           # Robust CI lower
    CI_upper = model$ci[3,2],           # Robust CI upper
    p_value = model$pv[3],              # Robust p-value
    Eff_N_total = sum(model$N_h),       # Total effective N
    Bandwidth = model$bws[1,1],         # Bandwidth
    stringsAsFactors = FALSE
  )
}

results_list <- lapply(names(models), function(name) {
  extract_rdd_results(models[[name]], name)
})

results_table <- do.call(rbind, results_list)
results_table_formatted <- results_table %>%
  mutate(
    Estimate = sprintf("%.4f", Estimate),
    CI_95 = sprintf("[%.4f, %.4f]", CI_lower, CI_upper),
    `p-value` = sprintf("%.4f", p_value),
    `Eff. N` = sprintf("%d", Eff_N_total),
    `Bandwidth` = sprintf("%.2f", Bandwidth)
  ) %>%
  select(Model, Estimate, CI_95, `p-value`, `Eff. N`, `Bandwidth`)

print(results_table_formatted)

stargazer(results_table_formatted, 
                    summary = FALSE, 
          rownames = FALSE, 
          type = "latex",
          out = "rdd_check_table.tex")
```

## c. graphs

```{r}
p_null <- ggplot(data_null, aes(x = z,
                       y = y1,       
                       color = factor(d))) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = TRUE) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  scale_x_continuous(
    breaks = seq(min(data$z, na.rm = TRUE),
                 max(data$z, na.rm = TRUE), by = 1)
  ) +
  scale_color_manual(
    values = c("0" = "red", "1" = "darkblue"),
    labels = c("Non-qualifier", "Qualifier"),
    name   = "Status"
  ) +
  labs(
    x     = "Points Margin",
    y     = "Nationalism (t+1)" 
  ) +
  theme_bw() +
  theme(legend.position = "bottom")

print(p_null)

ggsave("rdd_robustness_y1.png", p_null, width = 8, height = 6)
```

## d. Placebo checks

```{r}
library(rddensity)

rdd_dens <- rddensity(X=data$z,c=0)
summary(rdd_dens)

rdplotdensity(rdd_dens, data$z)
```


