---
title: "ETH: METHOD III DATABASE"
author: "Jüttner, D., Labarthe, S., Prastalo, N."
date: "2026"
output:
  pdf_document: default
  html_document: default
---
# CODEBOOK

## Identification 

* `match_id` - Unique numeric identifier for each match
* `tournament_id` - Unique identifier for the World Cup tournament
* `year` - Year of the World Cup tournament
* `group_name` - Name of the qualification group (e.g., "Group A")
* `group_stage` - Group stage identifier
* `stage_name` - Stage of the competition ("group stage")
* `match_date` - Date of the match (YYYY-MM-DD format)
* `country_name` - Name of the focal country/team
* `country_code` - ISO 3-letter country code (e.g., "FRA", "GER", "BRA")
* `team_id` - Unique numeric identifier for the focal team
* `opponent_name` - Name of the opposing country/team
* `opponent_id` - Unique numeric identifier for the opponent
* `opponent_code` - ISO 3-letter code of the opponent
* `venue` - Match venue from focal team's perspective ("home" or "away")

## Football variables

* `goals_for` - Number of goals scored by focal team
* `goals_against` - Number of goals conceded by focal team
* `goal_difference` - Goals for minus goals against (goal_difference = goals_for - goals_against)
* `win` - Binary indicator for win (1 = win, 0 = otherwise)
* `draw` - Binary indicator for draw (1 = draw, 0 = otherwise)
* `loss` - Binary indicator for loss (1 = loss, 0 = otherwise)
* `points` - Points earned in match (3 for win, 1 for draw, 0 for loss)
* `matches_played` - Total number of matches played by team in group stage
* `wins` - Total number of wins in group stage
* `draws` - Total number of draws in group stage
* `losses` - Total number of losses in group stage
* `rank` - Team's rank within the qualification group (1 = highest)
* `teams_qualifying` - Number of teams that qualify from each group (1 for year <= 1970, 2 otherwise)
* `qualified` - Binary indicator for World Cup qualification (1 = qualified, 0 = not qualified)
* `min_qualifier_points` - Minimum points among teams that qualified in the group
* `max_non_qualifier_points` - Maximum points among teams that did not qualify in the group
* `points_gap` - Difference between min_qualifier_points and max_non_qualifier_points
* `tie_at_cutoff` - Binary indicator for tie at qualification cutoff (1 = tie exists, 0 = otherwise)
* `treatment` - Binary treatment variable (1 = qualified, 0 = not qualified)
* `points_diff` - Running variable for RDD: points_gap if qualified, -points_gap if not qualified

## Military variables

* `bicc_milper` - Military Personnel Index (BICC, 1990-2023)
* `bicc_milexp` - Military Expenditure Index (BICC, 1990-2023)
* `bicc_gmi` - Global Militarization Index (BICC, 1990-2023)
* `wdi_expmil` - Military expenditure as percentage of GDP (World Bank, 1960-2024)

## Demographic variables

* `wdi_pop` - Total population (World Bank, 1960-2024)
* `wdi_pop_log` - Natural logarithm of total population
* `wdi_popurb` - Urban population (World Bank, 1960-2024)
* `wdi_birth` - Birth rate per 1,000 people (World Bank, 1960-2024)
* `wdi_death` - Death rate per 1,000 people (World Bank, 1960-2024)
* `wdi_mortinf` - Infant mortality per 1,000 live births (World Bank, 1960-2024)
* `wdi_lifexp` - Life expectancy at birth in years (World Bank, 1960-2023)
* `wpp_medianage` - Median age as of July 1 (UN World Population Prospects, 1950-2024)
* `wpp_sexratio` - Sex ratio (males per 100 females) (UN World Population Prospects, 1950-2024)

## Economic variables

* `wdi_export` - Exports of goods and services as percentage of GDP (World Bank, 1960-2024)
* `wdi_import` - Imports of goods and services as percentage of GDP (World Bank, 1960-2024)
* `wdi_gdpcapcon2015` - GDP per capita in constant 2015 US dollars (World Bank, 1960-2024)
* `wdi_gdpgr` - GDP growth rate (annual percentage) (World Bank, 1961-2024)
* `wdi_area` - Land area in square kilometers (World Bank, 1961-2023)

## Political variables

* `vdem_polyarchy` - Electoral democracy index (VDEM, 1946-2024)
* `ht_colonial` - Colonial origin indicator (HT, 1946-2025)
* `atop_number` - Number of alliances (ATOP, 1946-2025)
* `v2paseatshare` - Seat share of nationalist parties in parliament (VDEM)

## Notes

* `NA` - Missing values are present for some variables due to data availability limitations
* Time coverage varies by variable as indicated in parentheses above

# DATABASE REPLICATION
## Nationalism (share of seat in parliament)
```{r, eval = F}
#install.packages("remotes")  
remotes::install_github("vdeminstitute/vdemdata")

library(vdemdata)
library(dplyr)

vp <- vparty %>% 
  rename(country_code = country_text_id) %>% 
  select(country_code, year, v2paseatshare) %>% 
  group_by(country_code, year) %>%
  summarise(
    v2paseatshare = mean(v2paseatshare, na.rm = TRUE),
    .groups = "drop"
  )
```

## QOG
```{r, eval = F}
library(readr)

url_data <- "https://raw.githubusercontent.com/SimonLabarthe/Method-3/main/qog.rds"
qog <- read_rds(url_data) %>% 
  rename(country_name = cname, 
         country_code = ccodealp) %>% 
  select(
    country_name,
    country_code,
    year,

    bicc_milper,        # Military Personnel Index (i = 162, t=1990-2023)
    bicc_milexp,        # Military Expenditure Index (i = 162, t=1990-2023)
    bicc_gmi,           # Global Militarization Index (i = 162, t=1990-2023)
    wdi_expmil,         # Military expenditure % of GDP (i = 163, t=1960-2024))

    wdi_pop,            # Total Population (i = 192, t = 1960-2024)
    wdi_popurb,         # Urban Population (i = 192, t = 1960-2024)

    wdi_birth,          # Birth Rate (per 1,000) (i = 192, t = 1960-2024)
    wdi_death,          # Death Rate (per 1,000) (i = 192, t = 1960-2024)
    wdi_mortinf,        # Infant Mortality (per 1,000 live births) (i = 192, t = 1960-2024)
    wdi_lifexp,         # Life Expectancy at birth (per 1,000 live births) (i = 192, t = 1960-2023)

    wpp_medianage,      # Median Age as of 1 July (WPP) (i = 192, t = 1950-2024)
    wpp_sexratio,       # Sex Ratio: males per 100 females (WPP) (i = 192, t = 1950-2024)


    wdi_export,         # Exports of goods & services (% of GDP) (i = 173, t = 1960-2024)
    wdi_import,         # Imports of goods & services (% of GDP) (i = 173, t = 1960-2024)
    wdi_gdpcapcon2015,  # GDP per capita (constant 2015 US dollar) (i = 191, t = 1960-2024)
    wdi_gdpgr,          # GDP growth (annual %) (i = 191, t = 1961-2024)

    wdi_area,           # Land area in sq. km (i = 191, t = 1961-2023)

    vdem_polyarchy,     # Democracy level (i = 175, t = 1946-2024)
    ht_colonial,        # Colonial Origin (i = 198, t = 1946-2025)

    atop_number,        # Number of Alliances (i = 198, t = 1946-2025)
  ) %>% 
  mutate(
    wdi_pop_log = log(wdi_pop)
  )

summary(qog)
```

## FIFA World Cup
```{r, eval = F}
remotes::install_github("jfjelstul/worldcup")

library(tidyverse)
library(worldcup)
library(lubridate)

data("matches")
data("tournaments")
data("teams")
data("qualified_teams")

qualification_matches <- matches %>%
  filter(stage_name == "group stage") %>% 
  mutate(
    year = as.numeric(str_extract(tournament_name, "\\d{4}"))
  ) %>% 
  filter(year %% 2 == 0) # Only males

home_team_matches <- qualification_matches %>%
  select(
    tournament_id, 
    year, 
    match_id, 
    match_date,
    group_name, 
    group_stage, 
    stage_name,
    team_id = home_team_id,
    country_name = home_team_name,
    team_code = home_team_code,
    opponent_id = away_team_id,
    opponent_name = away_team_name,
    opponent_code = away_team_code,
    goals_for = home_team_score,
    goals_against = away_team_score,
    win = home_team_win,
    draw,
    loss = away_team_win
  ) %>%
  mutate(
    points = case_when(
      win == 1 ~ 3,
      draw == 1 ~ 1,
      TRUE ~ 0
    ),
    goal_difference = goals_for - goals_against,
    venue = "home"
  )

away_team_matches <- qualification_matches %>%
  select(
    tournament_id, 
    year, 
    match_id, 
    match_date,
    group_name, 
    group_stage, 
    stage_name,
    team_id = away_team_id,
    country_name = away_team_name,
    team_code = away_team_code,
    opponent_id = home_team_id,
    opponent_name = home_team_name,
    opponent_code = home_team_code,
    goals_for = away_team_score,
    goals_against = home_team_score,
    win = away_team_win,
    draw,
    loss = home_team_win
  ) %>%
  mutate(
    points = case_when(
      win == 1 ~ 3,
      draw == 1 ~ 1,
      TRUE ~ 0
    ),
    goal_difference = goals_for - goals_against,
    venue = "away"
  )

monadic_matches <- bind_rows(home_team_matches, 
                             away_team_matches) %>%
  arrange(year, 
          group_name, 
          country_name, 
          match_date)

qualification_standings <- monadic_matches %>%
  group_by(year, group_name, team_id, country_name, team_code) %>%
  summarise(
    matches_played = n(),
    wins = sum(win, na.rm = TRUE),
    draws = sum(draw, na.rm = TRUE),
    losses = sum(loss, na.rm = TRUE),
    goals_for = sum(goals_for, na.rm = TRUE),
    goals_against = sum(goals_against, na.rm = TRUE),
    goal_difference = sum(goal_difference, na.rm = TRUE),
    points = sum(points, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(year, group_name, desc(points), desc(goal_difference), desc(goals_for))

qualification_standings <- qualification_standings %>%
  group_by(year, group_name) %>%
  mutate(
    rank = row_number(),
    teams_qualifying = case_when(
      year <= 1970 ~ 1,  # Earlier tournaments had 1 qualifier per group
      year <= 2300 ~ 2,  # Later tournaments had 2
      TRUE ~ 2
    ),
    qualified = ifelse(rank <= teams_qualifying, 1, 0)
  ) %>%
  ungroup()

tight_races <- qualification_standings %>%
  rename(country_code = team_code) %>% 
  group_by(year, group_name) %>%
  mutate(
    min_qualifier_points = min(points[qualified == 1], na.rm = TRUE),
    max_non_qualifier_points = max(points[qualified == 0], na.rm = TRUE),
    points_gap = min_qualifier_points - max_non_qualifier_points,
    tie_at_cutoff = any(qualified == 1 & points == max_non_qualifier_points) |
                    any(qualified == 0 & points == min_qualifier_points)
  ) %>%
  ungroup() %>%
  filter(
    (qualified == 1 & points == min_qualifier_points) |
    (qualified == 0 & points == max_non_qualifier_points)
  ) %>%
  mutate(
    treatment = case_when(
      qualified == 1 ~ 1,
      qualified == 0 ~ 0
    ), 
    points_diff = ifelse(
      qualified == 1, points_gap, -points_gap
    )
  ) 
```

## Merge
```{r, eval = F}
library(countries)

data <- tight_races %>%
  left_join(qog, by = c("country_code", "year")) %>%
  left_join(vp, by = c("country_code", "year")) %>% 
  filter(!is.na(treatment), 
         points_diff != "Inf") 
```



