---
title: "ETH: METHOD III DATABASE"
date: "2026"
output:
  pdf_document: default
  html_document: default
---

# 1. CODEBOOK
* `iso3c` - ISO 3-letter country code
* `year` - Tournament year
* `v2exl_legitideolcr_0_1` - Nationalism (1 year lead)
* `v2exl_legitideolcr_0_3` - Nationalism (3 year lead)
* `treatment` - World Cup qualification treatment
* `points_gap` - Points gap between last qualifier and first non-qualifier
* `points_diff` - Assignment variable (points margin from cutoff)
* `wdi_expmil` - Military expenditure (% of GDP)
* `wdi_pop_log` - Total population (log)
* `wdi_popurb` - Urban population
* `wdi_lifexp` - Life expectancy at birth
* `wdi_export` - Exports of goods and services (% of GDP)
* `wdi_import` - Imports of goods and services (% of GDP)
* `wdi_gdpcapcon2015` - GDP per capita (constant 2015 US$)
* `wdi_gdpgr` - GDP growth (annual %)
* `wdi_area` - Land area (sq. km)
* `wpp_medianage` - Median age of population
* `wpp_sexratio` - Sex ratio (males per 100 females)
* `vdem_polyarchy` - Polyarchy (electoral democracy index)

# 2. DATABASE REPLICATION
```{r, eval = F}
## a. VDEM 
library(dplyr)
library(countrycode)
library(tidyr)
library(readr)

vdem <- read_rds("https://raw.githubusercontent.com/SimonLabarthe/Method-3/main/vdem.rds") %>%  
  mutate(v2exl_legitideolcr_0_1 = lead(v2exl_legitideolcr_0, 1), 
         v2exl_legitideolcr_0_3 = lead(v2exl_legitideolcr_0, 3)) 

vdem_clean <- vdem %>%
  mutate(
    country_name_std = case_when(
      country_name == "United States of America" ~ "United States",
      country_name == "Türkiye" ~ "Turkey",
      country_name == "Czechia" ~ "Czech Republic",
      country_name == "Burma/Myanmar" ~ "Myanmar",
      country_name == "Republic of Vietnam" ~ "Vietnam",
      country_name == "German Democratic Republic" ~ "Germany",
      TRUE ~ country_name
    ),
    iso3c = countrycode(country_name_std, 
                        "country.name", 
                        "iso3c",
                        warn = TRUE)
  ) %>%
  filter(!is.na(iso3c)) %>%
  select(-country_name_std) 

## b. QOG
url_data <- "https://raw.githubusercontent.com/SimonLabarthe/Method-3/main/qog.rds"
qog <- read_rds(url_data) %>% 
  rename(country_name = cname, 
         country_code = ccodealp) %>% 
  select(
    country_name,
    country_code,
    year,

    bicc_milper,        # Military Personnel Index (i = 162, t=1990-2023)
    bicc_milexp,        # Military Expenditure Index (i = 162, t=1990-2023)
    bicc_gmi,           # Global Militarization Index (i = 162, t=1990-2023)
    wdi_expmil,         # Military expenditure % of GDP (i = 163, t=1960-2024))

    wdi_pop,            # Total Population (i = 192, t = 1960-2024)
    wdi_popurb,         # Urban Population (i = 192, t = 1960-2024)

    wdi_birth,          # Birth Rate (per 1,000) (i = 192, t = 1960-2024)
    wdi_death,          # Death Rate (per 1,000) (i = 192, t = 1960-2024)
    wdi_mortinf,        # Infant Mortality (per 1,000 live births) (i = 192, t = 1960-2024)
    wdi_lifexp,         # Life Expectancy at birth (per 1,000 live births) (i = 192, t = 1960-2023)

    wpp_medianage,      # Median Age as of 1 July (WPP) (i = 192, t = 1950-2024)
    wpp_sexratio,       # Sex Ratio: males per 100 females (WPP) (i = 192, t = 1950-2024)


    wdi_export,         # Exports of goods & services (% of GDP) (i = 173, t = 1960-2024)
    wdi_import,         # Imports of goods & services (% of GDP) (i = 173, t = 1960-2024)
    wdi_gdpcapcon2015,  # GDP per capita (constant 2015 US dollar) (i = 191, t = 1960-2024)
    wdi_gdpgr,          # GDP growth (annual %) (i = 191, t = 1961-2024)

    wdi_area,           # Land area in sq. km (i = 191, t = 1961-2023)

    vdem_polyarchy,     # Democracy level (i = 175, t = 1946-2024)
    ht_colonial,        # Colonial Origin (i = 198, t = 1946-2025)

    atop_number,        # Number of Alliances (i = 198, t = 1946-2025)
  ) %>% 
  mutate(
    wdi_pop_log = log(wdi_pop)
  )

qog_clean <- qog %>%
  mutate(iso3c = countrycode(country_name, 
                             "country.name", 
                             "iso3c",
                             warn = TRUE)) %>%
  filter(!is.na(iso3c))

## c. FIFA World Cup
library(tidyverse)
library(worldcup)
library(lubridate)

data("matches")
data("tournaments")
data("teams")
data("qualified_teams")

qualification_matches <- matches %>%
  filter(stage_name == "group stage") %>% 
  mutate(
    year = as.numeric(str_extract(tournament_name, "\\d{4}"))
  ) %>% 
  filter(year %% 2 == 0) %>% # Only males 
  mutate(
    # Standardisation des noms de pays du foot
    country_name_std = case_when(
      country_name == "United States" ~ "United States",
      country_name == "Korea DPR" ~ "North Korea",
      country_name == "Korea Republic" ~ "South Korea",
      country_name == "Russia" ~ "Russia",
      country_name == "Iran" ~ "Iran",
      country_name == "Congo" ~ "Republic of Congo",
      country_name == "DR Congo" ~ "Democratic Republic of the Congo",
      country_name == "Ivory Coast" ~ "Côte d'Ivoire",
      country_name == "Bosnia-Herzegovina" ~ "Bosnia and Herzegovina",
      country_name == "England" ~ "United Kingdom",
      country_name == "Scotland" ~ "United Kingdom",
      country_name == "Wales" ~ "United Kingdom",
      country_name == "Northern Ireland" ~ "United Kingdom",
      country_name == "West Germany" ~ "Germany",
      country_name == "Czechoslovakia" ~ "Czech Republic",
      country_name == "Yugoslavia" ~ "Serbia",  # approximation
      country_name == "Soviet Union" ~ "Russia",  # approximation
      country_name == "Turkey" ~ "Turkey",
      TRUE ~ country_name
    ),
    iso3c = countrycode(country_name_std, 
                        "country.name", 
                        "iso3c",
                        custom_match = c(
                          "England" = "GBR",
                          "Scotland" = "GBR",
                          "Wales" = "GBR",
                          "Northern Ireland" = "GBR",
                          "West Germany" = "DEU",
                          "Czechoslovakia" = "CZE",
                          "Yugoslavia" = "SRB",
                          "Soviet Union" = "RUS"
                        ))
  ) %>%
  filter(!is.na(iso3c)) %>%
  select(-country_name_std)

home_team_matches <- qualification_matches %>%
  select(
    tournament_id, 
    iso3c,
    year, 
    match_id, 
    match_date,
    group_name, 
    group_stage, 
    stage_name,
    team_id = home_team_id,
    country_name = home_team_name,
    team_code = home_team_code,
    opponent_id = away_team_id,
    opponent_name = away_team_name,
    opponent_code = away_team_code,
    goals_for = home_team_score,
    goals_against = away_team_score,
    win = home_team_win,
    draw,
    loss = away_team_win
  ) %>%
  mutate(
    points = case_when(
      win == 1 ~ 3,
      draw == 1 ~ 1,
      TRUE ~ 0
    ),
    goal_difference = goals_for - goals_against,
    venue = "home"
  )

away_team_matches <- qualification_matches %>%
  select(
    tournament_id, 
    iso3c,
    year, 
    match_id, 
    match_date,
    group_name, 
    group_stage, 
    stage_name,
    team_id = away_team_id,
    country_name = away_team_name,
    team_code = away_team_code,
    opponent_id = home_team_id,
    opponent_name = home_team_name,
    opponent_code = home_team_code,
    goals_for = away_team_score,
    goals_against = home_team_score,
    win = away_team_win,
    draw,
    loss = home_team_win
  ) %>%
  mutate(
    points = case_when(
      win == 1 ~ 3,
      draw == 1 ~ 1,
      TRUE ~ 0
    ),
    goal_difference = goals_for - goals_against,
    venue = "away"
  )

monadic_matches <- bind_rows(home_team_matches, 
                             away_team_matches) %>%
  arrange(year, 
          group_name, 
          country_name, 
          match_date)

qualification_standings <- monadic_matches %>%
  group_by(year, group_name, team_id, country_name, team_code, iso3c) %>%
  summarise(
    matches_played = n(),
    wins = sum(win, na.rm = TRUE),
    draws = sum(draw, na.rm = TRUE),
    losses = sum(loss, na.rm = TRUE),
    goals_for = sum(goals_for, na.rm = TRUE),
    goals_against = sum(goals_against, na.rm = TRUE),
    goal_difference = sum(goal_difference, na.rm = TRUE),
    points = sum(points, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(year, group_name, desc(points), desc(goal_difference), desc(goals_for))

qualification_standings <- qualification_standings %>%
  group_by(year, group_name) %>%
  mutate(
    rank = row_number(),
    teams_qualifying = case_when(
      year <= 1970 ~ 1,  # Earlier tournaments had 1 qualifier per group
      year <= 2300 ~ 2,  # Later tournaments had 2
      TRUE ~ 2
    ),
    qualified = ifelse(rank <= teams_qualifying, 1, 0)
  ) %>%
  ungroup()

tight_races <- qualification_standings %>%
  rename(country_code = team_code) %>% 
  group_by(year, group_name) %>%
  mutate(
    min_qualifier_points = min(points[qualified == 1], na.rm = TRUE),
    max_non_qualifier_points = max(points[qualified == 0], na.rm = TRUE),
    points_gap = min_qualifier_points - max_non_qualifier_points,
    tie_at_cutoff = any(qualified == 1 & points == max_non_qualifier_points) |
                    any(qualified == 0 & points == min_qualifier_points)
  ) %>%
  ungroup() %>%
  filter(
    (qualified == 1 & points == min_qualifier_points) |
    (qualified == 0 & points == max_non_qualifier_points)
  ) %>%
  mutate(
    treatment = case_when(
      qualified == 1 ~ 1,
      qualified == 0 ~ 0
    ), 
    points_diff = ifelse(
      qualified == 1, points_gap, -points_gap
    ), 
    points_diff = ifelse(
      qualified == 1 & points_diff == 0, 0.01, points_diff), 
    points_diff = ifelse(
      qualified == 0 & points_diff == 0, -0.01, points_diff))

## d. Merge
data <- tight_races %>%
  left_join(qog_clean, by = c("iso3c", "year")) %>%
  left_join(vdem_clean, by = c("iso3c", "year")) %>% 
  filter(!is.na(treatment), 
         points_diff != "Inf") 

data <- data %>% 
  select(iso3c, year, v2exl_legitideolcr_0_1, v2exl_legitideolcr_0_3,  treatment, points_gap, points_diff, wdi_expmil, wdi_pop_log, wdi_popurb, wdi_lifexp, wdi_export, wdi_import, wdi_gdpcapcon2015, wdi_gdpgr, wdi_area, wpp_medianage, wpp_sexratio, vdem_polyarchy) %>% na.omit()

library(labelled)

var_labels <- c(
  "iso3c" = "ISO 3-letter country code",
  "year" = "Tournament year",
  "v2exl_legitideolcr_0_1" = "Nationalism (1 year lead)",
  "v2exl_legitideolcr_0_3" = "Nationalism (3 year lead)",
  "treatment" = "Treatment: World Cup qualification (1=qualified, 0=non-qualified)",
  "points_gap" = "Points gap between last qualifier and first non-qualifier",
  "points_diff" = "Assignment variable: points margin from cutoff (positive=qualified)",
  "wdi_expmil" = "Military expenditure (% of GDP)",
  "wdi_pop_log" = "Total population (log)",
  "wdi_popurb" = "Urban population",
  "wdi_lifexp" = "Life expectancy at birth",
  "wdi_export" = "Exports of goods and services (% of GDP)",
  "wdi_import" = "Imports of goods and services (% of GDP)",
  "wdi_gdpcapcon2015" = "GDP per capita (constant 2015 US$)",
  "wdi_gdpgr" = "GDP growth (annual %)",
  "wdi_area" = "Land area (sq. km)",
  "wpp_medianage" = "Median age of population",
  "wpp_sexratio" = "Sex ratio (males per 100 females)",
  "vdem_polyarchy" = "Polyarchy (electoral democracy index)"
)

data <- labelled::set_variable_labels(data, .labels = var_labels)

look_for(data)

saveRDS(data, "data.rds")
```

# APPENDIX
## a. Bertoli Replication file
```{r}
library(dplyr)
rcdata=read.csv("https://raw.githubusercontent.com/AndrewBertoli/World-Cup-Code/master/RegionalChampionshipData.csv",stringsAsFactors=FALSE) %>% 
  rename(country_name = Country, 
         year = Year)

vdem <- read_rds("https://raw.githubusercontent.com/SimonLabarthe/Method-3/main/vdem.rds") 

data <- rcdata %>% 
  left_join(vdem, by = c("country_name", "year"))

data <- data %>% 
  mutate(nationalism_1 = lead(v2exl_legitideolcr_0, 1))
```



